Bacterial Microbiome and Obesity Meta-analysis Sequence Processing Pipeline

### 2016-02-12 code/sequenceProcessing.Rmd
sequence processing updates

---
title: "Bacterial Microbiome and Obesity Meta-analysis Sequence Processing Pipeline"
author: "Marc Sze"
date: "January 26, 2016"
output: 
  html_document:
    keep_md: yes
bibliography: metaAnalysis.obesity.microbiome.bib
---
#### **Preamble**  
Each study is brought up one at a time.  This recount starts from time of download all the way to generation of the final shared file used for the meta-analysis in the study.  You will also need the latest version of mothur, SRA Toolkit, Metaphlan2, and Python 2.7 installed on your system.  There are also a few reference files that you will need.  The first is the [SILVA reference](http://www.mothur.org/wiki/Silva_reference_files) database and the second is the [RDP database](http://www.mothur.org/wiki/RDP_reference_files) of which we used release 102 and version 9 respectively.  Finally, any entered code is denoted will be surronded with ''.  As an example, ''print("Hello World)''.
<br>
<br>

##### **Baxter et al. 2016:**
This data set consisted of controls without colorectal cancer (CRC) and those with CRC.  The sequence reads of only the controls who did not have CRC were obtained directly from the first author since the data was generated within the Schloss lab.  However, the sequences can be found on the SRA under the accession number SRP062005. Only the raw 16S sequence reads for normal controls were processed  from the Sequence Read Archive (SRA) and subsequently used in the analysis.  So here we go....
<br>
<br>
I'm sure a smarter person could seperate the specific files that are normal only but I'm lazy and we had the storage space so I downloaded the entire data set.  
<br>
''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP062/SRP062005/''
<br>
''fastq-dump \*.sra''
<br><br>
The metadata can be found and downloaded by searching SRP062005 at the following site: http://www.ncbi.nlm.nih.gov/Traces/study/
<br><br>
The simple python program (Python 3.0) used to generate the .file file needed to make contigs can be found in this repository titled: *MSze.makefile.good.trial2.py* and the resulting file is titled *mothur.data.file*.  All you need to make this work is to have only the necessary fastq files in the respective folder.
<br><br>
Before getting started enter mothur and curate the silva.fasta for the v4 region with the following commands:
<br><br>
''pcr.seqs(fasta=silva.bacteria.fasta, start=11894, end=25319, keepdots=F)''
<br>
''system(mv silva.bacteria.pcr.fasta silva.v4.fasta)''
<br><br>

Next we utilized the batch mode process and can come back the next day to see everything completed.  Our command was the following:
<br><br>
''/share/scratch/schloss/mothur/mothur data1.batch''
<br>
<br>
If you thought this was difficult to get data just wait until later on in this pipeline to see just how crazy it can get trying to get the data to a useable point.
<br>
<br>

##### **Ross et al. 2015:**
This data set consisted of normal individuals from the Cameron Count Hispanic Cohort (CCHC).  The paper for this data set can be found [here](http://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-015-0072-y).  The project accession number in the SRA is SRP053023.  We can download the entire data set again with the code mentioned above but with one small substitution at the ending.  Here we go......
<br><br>
''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP053/SRP053023/''
<br>
<br>
Now since there was no sff file provided in this data set even though it is a 454 data set (I found this out when I tried to run the command -> sff-dump \*.sra).  We have to use the fastq function of the SRA Tool kit.
<br>
<br>
''fastq-dump \*.sra''
<br>
<br>
Because of this difference there are a few changes in this processing versus the standardized mothur 454 protocol.  First, instead of the sff.info() command we have to use the fastq.info() command.  Second, and perhaps more importantly, because we don't have the sff file we have to use the quality scores.  We used the author specified defaults of *qwindowaverage=25* and *qwindowsize=50*.  It should also be noted that we had *maxambig=5* and *maxhomop=8*.  Since the data come seperated already we do not need an oligo file and all we have to do is loop the mothur *trim.seqs* command. You can run a bash or shell program to get all of this done with either *#!/bin/sh* or *#!/bin/bash*.  So in a text editor:

''#!/bin/sh  
for sample in \$(ls \*sra)  
do  
fastq-dump \$sample  
done
<br><br>
for sample2 in \$(ls \*fastq)  
do  
/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=\$sample2)"  
rm \*logfile  
done
<br><br>
for TRIM in \$(ls ERR\*.fasta)  
do  
QFILE=\${TRIM//fasta/qual}  
/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=\$TRIM, qfile=\$QFILE, maxambig=5, maxhomop=8, qwindowaverage=25, qwindowsize=50, processors=8)"  
done  
rm \*.fastq''
<br><br>
The next lines of code are used to combine the fasta file, create a names file with the mothur unique command, and create a combined group file for subsequent processing.
<br><br>
''cat \*trim.fasta > combined.trim.fasta  
/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.trim.fasta)"  
grep -e '^>' combined.trim.fasta > test.fasta  
sed 's/>//' test.fasta > test2.fasta  
sed 's/\\./\_/g' test2.fasta > test3.fasta  
sed 's/_.\*//' test3.fasta > test4.fasta  
paste -d" " test2.fasta test4.fasta > combined.groups  
exit 0''
<br><br>
After this little program is run we follow the mothur 454 SOP as written and can use the following command to complete the processing using the put together *Ross.batch* file and running mothur.  One final difference from the standardized mothur approach that was made was during the screen.seqs command after alignment a hard cutoff *minlength=200* and *maxlength=250* was used.
<br><br>
''/share/scratch/schloss/mothur/mothur Ross.batch''
<br><br>
That completes the processing for this data set.  Simple so far I hope.  On to the next data set.
<br><br>

##### **Goodrich et al. 2014:**  
This is a second larger twin data set that builds on the Turnbaugh et al. findings.  The paper can be found [here](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC4255478/).  The individuals that participated in the study were part of the TwinsUK registry and is somewhat heavily skewed toward women.  The raw sequence data can be found on the SRA under ERP006339.  So using wget we can download the data with the following code:
<br><br>
''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/ERP/ERP006/ERP006339/''
<br><br>
 I started off with analyzing the data separately while sticking the the default mothur SOP for MiSeq processing.....
<br><br>
''#!/bin/sh  
\# Unpack sra files  
\# mkdir sra.complete  
for start in \$(ls \*sra)  
do  
fastq-dump \$start  
mv \$start sra.complete  
done
<br><br>
\# Unpack fastq files  
mkdir fastq.complete  
mkdir logfile.complete  
for info in \$(ls \*fastq)  
do  
/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=\$info)"  
mv \$info fastq.complete  
mv \*.logfile logfile.complete  
done
<br><br>
\# Screen the fasta file that came from the fastq  
mkdir fasta.complete  
for next in \$(ls \*fasta)  
do  
/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=\$next, maxambig=0, maxlength=275, processors=8)"  
mv \$next fasta.complete \# should store silva file in here and then move it working directory  
mv \*.logfile logfile.complete  
done
<br><br>
mv ~/twinsUK/fasta.complete/silva.v4.fasta ~/twinsUK  
\# Align the screened fasta file to V4 region of SILVA  
mkdir good.fasta.complete  
for sample2 in \$(ls \*good.fasta)  
do  
/share/scratch/schloss/mothur/mothur "#align.seqs(fasta=\$sample2, reference=silva.v4.fasta, processors=8)"  
mv \$sample2 good.fasta.complete  
mv \*logfile logfile.complete  
done
<br><br>
\#Screen the aligned sequences to specific location  
mkdir align.complete  
for sample3 in \$(ls \*.align)  
do  
/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=\$sample3, start=1968, end=11546, maxhomop=8, processors=8)"  
mv \$sample3 align.complete  
done
<br><br>
\# Filter the screened sequences to get a fasta file back  
mkdir align.good.complete  
for sample4 in \$(ls \*good.align)  
do  
/share/scratch/schloss/mothur/mothur "#filter.seqs(fasta=\$sample4, vertical=T, trump=., processors=8)"  
mv \$sample4 align.good.complete  
mv \*logfile logfile.complete  
done
<br><br>
\# Make the combined fasta file  
mkdir filter.fasta.complete  
cat \*filter.fasta > combined.fasta  
mv \*filter.fasta filter.fasta.complete
<br><br>
\# Make the combined group file  
grep -e '^>' combined.fasta > test.fasta  
sed 's/>//' test.fasta > test2.fasta  
sed 's/\\..\*\\$//' test2.fasta > test3.fasta  
paste -d"\\t" test2.fasta test3.fasta > combined.groups''
<br><br>
After this program was run I thought perhaps naively that I could simply run a set a batch file and be done with this data set.  Well not quite.  The sheer size and the way our server was configured meant that if I called much more than 130GB of RAM my job would run the risk of starting over regularly. So I did use the batch method but only upto *pre.cluster*.  After this the *chimera.uchime* step was just taking too long so I seperated everything again and did the *chimera.uchime* separately.  This also meant having to redo the unique seqs step.  The code for this is below:
<br><br>
''/share/scratch/schloss/mothur/mothur Goodrich.batch''
<br><br>
So now the biggest challenge that I have found so far is getting the data into a distance matrix not based on the closed-reference approach.  The file is just too big for cluster.split so I'm going to have to find an alternative way.

@@

##### **Escobar et al. 2014:**
This data set consisted of normal individuals from Medellin, Colombia South America.  The paper for this data set can be found [here](http://bmcmicrobiol.biomedcentral.com/articles/10.1186/s12866-014-0311-6).  The project accession number in the ENA is ERP003466.  We can download the entire data set again with the code mentioned above but with a few substitutions versus the previous two studies.  Here we go......
<br><br>
''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/ERP/ERP003/ERP003466/''
<br>
<br>
Now just like the previous data set there was no sff file provided in this data set even though it is a 454 data set (I found this out, again, when I tried to run the command -> sff-dump \*.sra).  We have to use the fastq function of the SRA Tool kit.
<br>
<br>
''fastq-dump \*.sra''
<br>
<br>
Because of this difference there are a few changes in this processing versus the standardized mothur 454 protocol.  First, instead of the sff.info() command we have to use the fastq.info() command.  Second, and perhaps more importantly, because we don't have the sff file we have to use the quality scores.  Unlike the Ross, et al study we used the mothur specified defaults of *qwindowaverage=35* and *qwindowsize=50*.  It should also be noted that we had *maxambig=0* and *maxhomop=8*.  Since the data come seperated already we do not need an oligo file and all we have to do is loop the mothur *trim.seqs* command. You can run a bash or shell program to get all of this done with either *#!/bin/sh* or *#!/bin/bash*.  So in a text editor:

''#!/bin/sh  
for sample in \$(ls \*sra)  
do  
fastq-dump \$sample  
done
<br><br>
for sample2 in \$(ls \*fastq)  
do  
/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=\$sample2)"  
rm \*logfile  
done
<br><br>
for TRIM in \$(ls SRR\*.fasta)  
do  
QFILE=\${TRIM//fasta/qual}  
/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=\$TRIM, qfile=\$QFILE, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50, processors=8)"  
done  
rm \*.fastq''
<br><br>
The next lines of code are used to combine the fasta file, create a names file with the mothur unique command, and create a combined group file for subsequent processing.
<br><br>
''cat \*trim.fasta > combined.trim.fasta  
/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.trim.fasta)"  
grep -e '^>' combined.trim.fasta > test.fasta  
sed 's/>//' test.fasta > test2.fasta  
sed 's/\\./\_/g' test2.fasta > test3.fasta  
sed 's/_.\*//' test3.fasta > test4.fasta  
paste -d" " test2.fasta test4.fasta > combined.groups  
exit 0''
<br><br>
After this little program is run we follow the mothur 454 SOP as written and can use the following command to complete the processing using the put together *Escobar.batch* file and running mothur.  Again like the Ross et al data set one final difference from the standardized mothur approach that was made was during the screen.seqs command after alignment a hard cutoff *minlength=200* and *maxlength=250* was used.
<br><br>
''/share/scratch/schloss/mothur/mothur Escobar.batch''
<br><br>
That completes the processing for this data set.  So far so good, no major problems.  On to the next data set.
<br><br>

##### **Zupancic et al. 2012:**
This data set consisted of normal individuals from an Amish community living in Lancaster County, Pennsylvania.  The paper for this data set can be found [here](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.00430526).  The project accession number in the SRA is SRP002465.  We can download the entire data set again with the code mentioned above but with a few substitutions.  Here we go......
''wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP002/SRP002465/''

@@

''#!/bin/sh
for sample in \$(ls \*sra)  
do  
sff-dump \$sample  

@@

for sample2 in \$(ls \*sff)  
do  
/share/scratch/schloss/mothur/mothur "#sffinfo(sff=\$sample2, flow=T)"  
rm \*logfile  

@@

for FLOW in \$(ls \*flow)  
do  
/share/scratch/schloss/mothur/mothur "#trim.flows(flow=\$FLOW, processors=8)"  
rm \*logfile  
for SHHH in \$(ls \*files)  
do  
/share/scratch/schloss/mothur/mothur "#shhh.flows(file=\$SHHH, processors=8)"  
rm \*logfile  
for TRIMS in \$(ls \*shhh.fasta)  
do  
NAMES=\${TRIMS//fasta/names}  
/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=\$TRIMS, name=\$NAMES, maxhomop=8, maxambig=0, minlength=200, processors=8)"  
rm \*logfile  
exit 0''
Now, I had a few extra lines that combined the finalized files together but when I went to analyze these they gave some weird errors.  So I needed to make a second program to sort through these errors and analyze the data separately still.  One thing I noticed was that by using the standard workflow a large number of files ended up with 0 or very few sequences passing the quality filters.  The other problem which I delayed dealing with in this data set was that many samples were done sequenced more than once. This in itself is not a problem but the sequence ID associated with these sequences from different files were the exact same and so were carbon copies.  This wreaked havoc when I tried to combined the data together.  But I'll deal with that problem later.  I'm just going to continue on with the pipeline with the file seperated.  I will however, have to deal with this before the preclustering step.
So here is the second program used to keep processing the sequences separately rather than as one combined fasta file.....
''#!/bin/sh  
\#Align the fasta file to the SILVA database and create directories to move files into    
mkdir good.fasta.complete  
mkdir reference.files  
mkdir logifle.complete  
mv silva* reference files  
mv trainset9* reference files  
for sample2 in \$(ls \*.fasta)  
do  
/share/scratch/schloss/mothur/mothur "#align.seqs(fasta=\$sample2, reference=reference.files/silva.bacteria.fasta, flip=t, processors=8)"  
mv \$sample2 good.fasta.complete  
mv \*logfile logfile.complete  
\#Screen the align sequences  
mkdir align.complete  
for sample3 in \$(ls \*.align)  
do  
/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=\$sample3, end=6453, optimize=start, criteria=95, processors=8)"  
mv \$sample3 align.complete
mv \*logfile logfile.complete  
\#filter the screen sequences  
mkdir align.good.complete  
for sample4 in \$(ls \*good.align)  
do  
/share/scratch/schloss/mothur/mothur "#filter.seqs(fasta=\$sample4, vertical=T, trump=., processors=8)"  
mv \$sample4 align.good.complete  
mv \*logfile logfile.complete  
exit 0''
Right, now I need to pre cluster the data and I can't get away with analyzing each file separately anymore.  However, remember from up above there is something wierd going on in that the unique sequence ID are duplicated for many samples and there were a whole bunch of samples that did not survive quality filtering.  Since some samples had two samples others had 6 samples to be fair I decided to take the first sample from an individual that had sequences in them (so greater than 0 MB).  Now, this was done when I was still truly a novice in using shell scripting so I did this the "hard" way (i.e. the following method is pretty manual with very little automation so someone can make this more automated to improve on this process).  Essentially if a sample had sequence in it the indivdiual study ID was assigned to the fasta file and stored in a new directory.  Those that either had 0 MB after this we carried over and the next sequence ID file was used.  This was done until all samples had an indivdiual ID.  In my novice approach the only thing I automated was the original copying of the sequence ID file to a individual study ID file.  I then visually checked the size and manually sorted the files into directories.  From this I created a new list to run through.  The lists that were used were *combined2*, *combined3*, *combined4*, *combined5*, *combined6*, *combined7*, and *combined8* and can be found in this same repository.

>>>>

---
title: "Bacterial Microbiome and Obesity Meta-analysis Sequence Processing Pipeline"
author: "Marc Sze"
date: "January 26, 2016"
output: 
  html_document:
    keep_md: yes
bibliography: metaAnalysis.obesity.microbiome.bib
---
#### **Preamble**  
Each study is brought up one at a time.  This recount starts from time of download all the way to generation of the final shared file used for the meta-analysis in the study.  You will also need the latest version of mothur, SRA Toolkit, Metaphlan2, Python 2.7, and Python 3+ installed on your system.  There are also a few reference files that you will need.  The first is the [SILVA reference](http://www.mothur.org/wiki/Silva_reference_files) database and the second is the [RDP database](http://www.mothur.org/wiki/RDP_reference_files) of which we used release 102 and version 9 respectively.  In general, we tried to keep close to the Schloss SOP for processsing sequence reads.  However, where this was not possible (mostly due to a high loss of reads resulting in a high loss of samples) we defaulted to the processing used by the published authors.  Where it was not possible to use a *de novo* approach (mostly due to high numbers of unique sequences) and the authors used a closed-reference approach we used the RDP database instead of greengenes.

<br>
<br>

##### **Baxter et al. 2016:**
This data set consisted of controls without colorectal cancer (CRC) and those with CRC.  The sequence reads of only the controls who did not have CRC were obtained directly from the first author since the data was generated within the Schloss lab.  However, the sequences can be found on the SRA under the accession number SRP062005. Only the raw 16S sequence reads for normal controls were processed  from the Sequence Read Archive (SRA) and subsequently used in the analysis.  So here we go....
<br>
<br>
I'm sure a smarter person could seperate the specific files that are normal only but I'm lazy and we had the storage space so I downloaded the entire data set.  
<br>
```{r, engine='bash', eval=FALSE}
wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP062/SRP062005/

fastq-dump \*.sra
```
<br><br>
The metadata can be found and downloaded by searching SRP062005 at the following site: http://www.ncbi.nlm.nih.gov/Traces/study/
<br><br>
The simple python program (Python 3.0) used to generate the .file file needed to make contigs can be found in this repository titled: *MSze.makefile.good.trial2.py* and the resulting file is titled *mothur.data.file*.  All you need to make this work is to have only the necessary fastq files in the respective folder.
<br><br>
Before getting started enter mothur and curate the silva.fasta for the v4 region with the following commands:

```{r, engine='bash', eval=FALSE}
pcr.seqs(fasta=silva.bacteria.fasta, start=11894, end=25319, keepdots=F)
system(mv silva.bacteria.pcr.fasta silva.v4.fasta)
```

Next we utilized the batch mode process and can come back the next day to see everything completed.  Our command was the following:

```{r, engine='bash', eval=FALSE}
/share/scratch/schloss/mothur/mothur data1.batch
```

If you thought this was difficult to get data just wait until later on in this pipeline to see just how crazy it can get trying to get the data to a useable point.
<br>
<br>

##### **Ross et al. 2015:**
This data set consisted of normal individuals from the Cameron Count Hispanic Cohort (CCHC).  The paper for this data set can be found [here](http://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-015-0072-y).  The project accession number in the SRA is SRP053023.  We can download the entire data set again with the code mentioned above but with one small substitution at the ending.  Here we go......

```{r, engine='bash', eval=FALSE}
wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP053/SRP053023/
```

Now since there was no sff file provided in this data set even though it is a 454 data set (I found this out when I tried to run the command -> sff-dump \*.sra).  We have to use the fastq function of the SRA Tool kit.
<br>
```{r, engine='bash', eval=FALSE}
fastq-dump \*.sra
```
<br>
Because of this difference there are a few changes in this processing versus the standardized mothur 454 protocol.  First, instead of the sff.info() command we have to use the fastq.info() command.  Second, and perhaps more importantly, because we don't have the sff file we have to use the quality scores.  We used the author specified defaults of *qwindowaverage=25* and *qwindowsize=50*.  It should also be noted that we had *maxambig=5* and *maxhomop=8*.  Since the data come seperated already we do not need an oligo file and all we have to do is loop the mothur *trim.seqs* command. You can run a bash or shell program to get all of this done with either *#!/bin/sh* or *#!/bin/bash*.  So in a text editor:

```{r, engine='sh', eval=FALSE}
#!/bin/sh  
for sample in $(ls *sra)  
do  
fastq-dump $sample  
done

for sample2 in $(ls *fastq)  
do  
/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$sample2)"  
rm *logfile  
done

for TRIM in $(ls ERR*.fasta)  
do  
QFILE=${TRIM//fasta/qual}  
/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$TRIM, qfile=$QFILE, maxambig=5, maxhomop=8, qwindowaverage=25, qwindowsize=50, processors=8)"  
done

rm *.fastq
```

The next lines of code are used to combine the fasta file, create a names file with the mothur unique command, and create a combined group file for subsequent processing.

```{r, engine='bash', eval=FALSE}
cat \*trim.fasta > combined.trim.fasta  
/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.trim.fasta)"

grep -e '^>' combined.trim.fasta > test.fasta  
sed 's/>//' test.fasta > test2.fasta  
sed 's/\./_/g' test2.fasta > test3.fasta  
sed 's/_.*//' test3.fasta > test4.fasta  
paste -d" " test2.fasta test4.fasta > combined.groups

exit 0
```

After this little program is run we follow the mothur 454 SOP as written and can use the following command to complete the processing using the put together *Ross.batch* file and running mothur.  One final difference from the standardized mothur approach that was made was during the screen.seqs command after alignment a hard cutoff *minlength=200* and *maxlength=250* was used.

```{r, engine='bash', eval=FALSE}
/share/scratch/schloss/mothur/mothur Ross.batch
```

That completes the processing for this data set.  Simple so far I hope.  On to the next data set.
<br><br>

##### **Goodrich et al. 2014:**  
This is a second larger twin data set that builds on the Turnbaugh et al. findings.  The paper can be found [here](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC4255478/).  The individuals that participated in the study were part of the TwinsUK registry and is somewhat heavily skewed toward women.  The raw sequence data can be found on the SRA under ERP006339.  So using wget we can download the data with the following code:

```{r, engine='bash', eval=FALSE}
wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/ERP/ERP006/ERP006339/

```

I started off with analyzing the data separately while sticking the the default mothur SOP for MiSeq processing.....

```{r, engine='sh', eval=FALSE}
#!/bin/sh  
# Unpack sra files  
# mkdir sra.complete  
for start in \$(ls \*sra)  
do  
fastq-dump $start  
mv $start sra.complete  
done

# Unpack fastq files  
mkdir fastq.complete  
mkdir logfile.complete  
for info in $(ls *fastq)  
do  
/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$info)"  
mv $info fastq.complete  
mv *.logfile logfile.complete  
done

# Screen the fasta file that came from the fastq  
mkdir fasta.complete  
for next in $(ls *fasta)  
do  
/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=$next, maxambig=0, maxlength=275, processors=8)"  
mv $next fasta.complete # should store silva file in here and then move it to working directory  
mv *.logfile logfile.complete  
done

mv ~/twinsUK/fasta.complete/silva.v4.fasta ~/twinsUK  
# Align the screened fasta file to V4 region of SILVA  
mkdir good.fasta.complete  
for sample2 in $(ls *good.fasta)  
do  
/share/scratch/schloss/mothur/mothur "#align.seqs(fasta=$sample2, reference=silva.v4.fasta, processors=8)"  
mv $sample2 good.fasta.complete  
mv *logfile logfile.complete  
done

#Screen the aligned sequences to specific location  
mkdir align.complete  
for sample3 in $(ls *.align)  
do  
/share/scratch/schloss/mothur/mothur "#screen.seqs(fasta=$sample3, start=1968, end=11546, maxhomop=8, processors=8)"  
mv $sample3 align.complete  
done

# Filter the screened sequences to get a fasta file back  
mkdir align.good.complete  
for sample4 in $(ls *good.align)  
do  
/share/scratch/schloss/mothur/mothur "#filter.seqs(fasta=$sample4, vertical=T, trump=., processors=8)"  
mv $sample4 align.good.complete  
mv *logfile logfile.complete  
done

# Make the combined fasta file  
mkdir filter.fasta.complete  
cat *filter.fasta > combined.fasta  
mv *filter.fasta filter.fasta.complete

# Make the combined group file  
grep -e '^>' combined.fasta > test.fasta  
sed 's/>//' test.fasta > test2.fasta  
sed 's/\..*$//' test2.fasta > test3.fasta  
paste -d"\t" test2.fasta test3.fasta > combined.groups

```

After this program was run I thought perhaps naively that I could simply run a set a batch file and be done with this data set.  Well not quite.  The sheer size and the way our server was configured meant that if I called much more than 130GB of RAM my job would run the risk of starting over regularly. So I did use the batch method but only upto *pre.cluster*.  After this the *chimera.uchime* step was just taking too long so I seperated everything again and did the *chimera.uchime* separately.  This also meant having to redo the unique seqs step.  The code for this is below:

```{r, engine='bash', eval=FALSE}
/share/scratch/schloss/mothur/mothur Goodrich.batch
```

So now the biggest challenge that I have found so far is getting the data into a distance matrix not based on the closed-reference approach.  The file is just too big for cluster.split so we have had to use the closed-reference approach for this data set.

```{r, engine='bash', eval=FALSE}
/share/scratch/schloss/mothur/mothur Goodrich2.batch
```

@@

##### **Escobar et al. 2014:**
This data set consisted of normal individuals from Medellin, Colombia South America.  The paper for this data set can be found [here](http://bmcmicrobiol.biomedcentral.com/articles/10.1186/s12866-014-0311-6).  The project accession number in the ENA is ERP003466.  We can download the entire data set again with the code mentioned above but with a few substitutions versus the previous two studies.  Here we go......

```{r, engine='bash', eval=FALSE}
wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/ERP/ERP003/ERP003466/
```

Now just like the previous data set there was no sff file provided in this data set even though it is a 454 data set (I found this out, again, when I tried to run the command -> sff-dump \*.sra).  We have to use the fastq function of the SRA Tool kit.

```{r, engine='bash', eval=FALSE}
fastq-dump *.sra
```

Because of this difference there are a few changes in this processing versus the standardized mothur 454 protocol.  First, instead of the sff.info() command we have to use the fastq.info() command.  Second, and perhaps more importantly, because we don't have the sff file we have to use the quality scores.  Unlike the Ross, et al study we used the mothur specified defaults of *qwindowaverage=35* and *qwindowsize=50*.  It should also be noted that we had *maxambig=0* and *maxhomop=8*.  Since the data come seperated already we do not need an oligo file and all we have to do is loop the mothur *trim.seqs* command. You can run a bash or shell program to get all of this done with either *#!/bin/sh* or *#!/bin/bash*.  So in a text editor:

```{r, engine='sh', eval=FALSE}
#!/bin/sh  
for sample in $(ls *sra)  
do  
fastq-dump $sample  
done
<br><br>
for sample2 in $(ls *fastq)  
do  
/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$sample2)"  
rm *logfile  
done
<br><br>
for TRIM in $(ls SRR*.fasta)  
do  
QFILE=\${TRIM//fasta/qual}  
/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$TRIM, qfile=$QFILE, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50, processors=8)"  
done  
rm *.fastq
```

The next lines of code are used to combine the fasta file, create a names file with the mothur unique command, and create a combined group file for subsequent processing.

```{r, engine='bash', eval=FALSE}
cat *trim.fasta > combined.trim.fasta

/share/scratch/schloss/mothur/mothur "#unique.seqs(fasta=combined.trim.fasta)"

grep -e '^>' combined.trim.fasta > test.fasta  
sed 's/>//' test.fasta > test2.fasta  
sed 's/\./_/g' test2.fasta > test3.fasta  
sed 's/_.*//' test3.fasta > test4.fasta  
paste -d" " test2.fasta test4.fasta > combined.groups  
exit 0
```

After this little program is run we follow the mothur 454 SOP as written and can use the following command to complete the processing using the put together *Escobar.batch* file and running mothur.  Again like the Ross et al data set one final difference from the standardized mothur approach that was made was during the screen.seqs command after alignment a hard cutoff *minlength=200* and *maxlength=250* was used.

```{r, engine='bash', eval=FALSE}
/share/scratch/schloss/mothur/mothur Escobar.batch
```

That completes the processing for this data set.  So far so good, no major problems.  On to the next data set.
<br><br>

##### **Zupancic et al. 2012:**
This data set consisted of normal individuals from an Amish community living in Lancaster County, Pennsylvania.  The paper for this data set can be found [here](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.00430526).  The project accession number in the SRA is SRP002465.  We can download the entire data set again with the code mentioned above but with a few substitutions.  Here we go......
```{r, engine='bash', eval=FALSE}
wget -r -np -nd -k ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/SRP002/SRP002465/
```

@@

```{r, engine='bash', eval=FALSE}
#!/bin/bash

# Make a directory to store the reference files
mkdir reference.files
mv silva.bacteria.fasta reference.files
mv trainset9_032012.pds.tax reference.files
mv trainset9_032012.pds.fasta reference.files

# Read in the sra and make a fastq instead of a sff.
mkdir completed.redo.sra
mkdir logfile.completed

for sample in $(ls *sra)
do
	fastq-dump $sample
	mv $sample completed.redo.sra
	mv *logfile logfile.completed

@@

# Get the quality file and fasta information from the fastq
mkdir completed.fastq

for sample2 in $(ls *fastq)
do
	/share/scratch/schloss/mothur/mothur "#fastq.info(fastq=$sample2)"	
	mv $sample2 completed.fastq
	mv *logfile logfile.completed

@@

# Trim the fastas according to quality 
mkdir first.trim.completed
for Trimming in $(ls *fasta)
do
	Quality=${Trimming//fasta/qual}
	/share/scratch/schloss/mothur/mothur "#trim.seqs(fasta=$Trimming, qfile=$Quality, maxambig=0, maxhomop=8, qwindowaverage=35, qwindowsize=50)"
	mv $Trimming first.trim.completed
	mv $Quality first.trim.completed
	mv *logfile logfile.completed
done

mkdir scrap.fasta.completed
mkdir scrap.qual.completed
mkdir trim.qual.completed
mv *scrap.qual scrap.qual.completed
mv * scrap.fasta.completed
mv *trim.qual trim.qual.completed
```

Now I realized moving through this data set that they had SRR's for an individual sample so I decided to combine all these SRR's together based on their sample ID.  I used a python script that would make this possible.  It takes to columns seperated by a tab and looks for common names from one column and inserts a user defined command at the beginning as well as any special symbols like /* or .fasta to the end of each column as needed.  If you want to use the program it is found in my personal repository [here](https://github.com/marcsze/pythonPrograms) called RepeatEdit.py.  At any rate if you just want to copy and paste the command (it is long) you can see it below.  The output of the RepeatEdit.py program will basically spit this out for you.

```{r, engine='bash', eval=FALSE}
#!/bin/bash

cat SRR050640* SRR050641* > p0_s2.fasta
cat SRR050550* SRR050551* SRR050806* SRR050807* > p0_s1.fasta
cat SRR050558* SRR050559* SRR050814* SRR050815* SRR050732* SRR050733* > p10_s1.fasta
cat SRR050808* SRR050809* SRR050552* SRR050553* > p100_s1.fasta
cat SRR050724* SRR050725* > p103_s1.fasta
cat SRR050452* SRR050453* SRR050650* SRR050651* > p115_s1.fasta
cat SRR050456* SRR050457* > p118_s1.fasta
cat SRR050460* SRR050461* > p120_s2.fasta
cat SRR050556* SRR050557* SRR050812* SRR050813* > p108_s1.fasta
cat SRR050560* SRR050561* SRR050816* SRR050817* > p111_s1.fasta
cat SRR050642* SRR050643* > p106_s2.fasta
cat SRR050644* SRR050645* > p112_s1.fasta
cat SRR050648* SRR050649* > p114_s1.fasta
cat SRR050652* SRR050653* > p119_s1.fasta
cat SRR050728* SRR050729* > p105_s1.fasta
cat SRR050730* SRR050731* > p105_s2.fasta
cat SRR050554* SRR050555* SRR050810* SRR050811* > p106_s1.fasta
cat SRR050458* SRR050459* > p120_s1.fasta
cat SRR050464* SRR050465* > p124_s1.fasta
cat SRR050466* SRR050467* > p125_s1.fasta
cat SRR050470* SRR050471* > p129_s1.fasta
cat SRR050472* SRR050473* > p130_s1.fasta
cat SRR050476* SRR050477* > p133_s1.fasta
cat SRR050478* SRR050479* > p134_s1.fasta
cat SRR050482* SRR050483* > p136_s1.fasta
cat SRR050484* SRR050485* > p137_s1.fasta
cat SRR050488* SRR050489* > p140_s1.fasta
cat SRR050490* SRR050491* > p142_s1.fasta
cat SRR050494* SRR050495* > p148_s1.fasta
cat SRR050496* SRR050497* > p149_s1.fasta
cat SRR050500* SRR050501* > p151_s1.fasta
cat SRR050502* SRR050503* > p152_s1.fasta
cat SRR050522* SRR050523* > p36_s1.fasta
cat SRR050524* SRR050525* > p39_s2.fasta
cat SRR050528* SRR050529* > p42_s1.fasta
cat SRR050530* SRR050531* SRR050682* SRR050683* SRR095127* SRR095128* > p44_s2.fasta
cat SRR050564* SRR050565* SRR050820* SRR050821* > p14_s1.fasta
cat SRR050568* SRR050569* SRR050824* SRR050825* > p17_s1.fasta
